The hedging hyperprior for the normal distribution

### 2023-03-09 text/23/exponential-hedging-hyperprior.Rmd
Converted normal-hedging-hyperprior to use precisions

```{r}
tau_grid <- seq(0, 1, length=60)
theta_grid <- seq(0, x_max, length=60)

gstar <- function(theta, tau) {g(theta, tau*(a0-1)+1, tau*(b0-1)+1)}
density <- outer(theta_grid, tau_grid, gstar)
op <- par(bg = "white")
z_max <- max(density, na.rm=TRUE)
persp(
  theta_grid, 
  tau_grid, 
  density, 
  zlim=c(0, 1.1*z_max),
  xlab="theta",
  ylab="tau",
  theta = 135, 
  phi = 30, 
  expand=0.5,
  ticktype="detailed")
```

It is hard to see at this scale, but the prior distribution is higher for small values of $\tau$ when $\theta4 is very large or very small. This becomes more obvious if you remove the middle of the surface and rescale the Z-axis.

```{r}
density <- outer(theta_grid, tau_grid, gstar)
density[11:40, ] <- NA
op <- par(bg = "white")
z_max <- max(density, na.rm=TRUE)
persp(
  theta_grid, 
  tau_grid, 
  density, 
  zlim=c(0, 1.1*z_max),
  xlab="theta",
  ylab="tau",
  theta = 135, 
  phi = 30, 
  expand=0.5,
  ticktype="detailed")
```

The likelihood function is constant across all values of $\tau$, so its surface looks like the following.

```{r}
tau_grid <- seq(0, 1, length=60)
theta_grid <- seq(0, x_max, length=60)

fstar <- function(theta, tau) {f(xbar, theta)^n}
likelihood <- outer(theta_grid, tau_grid, fstar)
z_max <- max(likelihood, na.rm=TRUE)
persp(
  theta_grid, 
  tau_grid, 
  likelihood, 
  zlim=c(0, 1.1*z_max),
  xlab="theta",
  ylab="tau",
  theta = 135, 
  phi = 30, 
  expand=0.5,
  ticktype="detailed")
```

@@

```{r}
tau_grid <- seq(0, 1, length=60)
theta_grid <- seq(0, x_max, length=60)

hstar <- function(theta, tau) {g(theta, tau*(a0-1)+1, tau*(b0-1)+1)*f(xbar, theta)^n}
density <- outer(theta_grid, tau_grid, hstar)
z_max <- max(density, na.rm=TRUE)
persp(
  theta_grid, 
  tau_grid, 
  density, 
  zlim=c(0, 1.1*z_max),
  xlab="theta",
  ylab="tau",
  theta = 135, 
  phi = 30, 
  expand=0.5,
  ticktype="detailed")
```

@@

```{r}
density <- density / sum(density, na.rm=TRUE)
density %>%
  as.vector %>%
  sort %>%
  rev -> sorted_density
cumulative_p <- 0
for (p in sorted_density) {
  cumulative_p <- p + cumulative_p
  density[density==p] <- cumulative_p
}
contour(
  theta_grid, 
  tau_grid, 
  density,
  labcex=1,
  levels=c(10, 25, 50, 75, 90)/100,
  xlab="theta",
  ylab="tau")
```

You can see that the hedging hyperprior did its job. The most probable values of $\tau$ are small and the prior mean (`r b0/(a0-1)`) is largely ignored in favor of the mean of the data (`r xbar`).

>>>>

```{r}
lambda_grid <- seq(0, 1, length=60)
theta_grid <- seq(0, x_max, length=60)

gstar <- function(theta, lambda) {g(theta, lambda*(a0-1)+1, lambda*(b0-1)+1)}
density <- outer(theta_grid, lambda_grid, gstar)
op <- par(bg = "white")
z_max <- max(density, na.rm=TRUE)
persp(
  theta_grid, 
  lambda_grid, 
  density, 
  zlim=c(0, 1.1*z_max),
  xlab="theta",
  ylab="lambda",
  theta = 135, 
  phi = 30, 
  expand=0.5,
  ticktype="detailed")
```

It is hard to see at this scale, but the prior distribution is higher for small values of $\lambda$ when $\theta$ is very large. This becomes more obvious if you remove the middle of the surface and rescale the Z-axis. There is a similar pattern on the far side of the graph. The prior trends towards a higher prior for small values of $\lambda$ when $\theta$ is very small.

This illustrates how the hedging hyperprior does its magic. It puts more probability on large values of $\lambda$ (gives closer to full weight for the informative prior) when you observe data values close to the prior mean. It puts more probability on small values of $\lambda$ (gives much less weight to the informative prior) when the data values are much larger or much smaller than the prior mean.

```{r}
density <- outer(theta_grid, lambda_grid, gstar)
# density[12:32, ] <- NA
# density[11:40, ] <- NA
density[density>0.3] <- 0.3
op <- par(bg = "white")
z_max <- max(density, na.rm=TRUE)
persp(
  theta_grid, 
  lambda_grid, 
  density, 
  zlim=c(0, 1.1*z_max),
  xlab="theta",
  ylab="lambda",
  theta = 135, 
  phi = 30, 
  expand=0.5,
  ticktype="detailed")
```

The likelihood function is constant across all values of $\lambda$, so its surface looks like the following.

```{r}
lambda_grid <- seq(0, 1, length=60)
theta_grid <- seq(0, x_max, length=60)

fstar <- function(theta, lambda) {f(xbar, theta)^n}
likelihood <- outer(theta_grid, lambda_grid, fstar)
z_max <- max(likelihood, na.rm=TRUE)
persp(
  theta_grid, 
  lambda_grid, 
  likelihood, 
  zlim=c(0, 1.1*z_max),
  xlab="theta",
  ylab="lambda",
  theta = 135, 
  phi = 30, 
  expand=0.5,
  ticktype="detailed")
```

@@

```{r}
lambda_grid <- seq(0, 1, length=60)
theta_grid <- seq(0, x_max, length=60)

hstar <- function(theta, lambda) {g(theta, lambda*(a0-1)+1, lambda*(b0-1)+1)*f(xbar, theta)^n}
density <- outer(theta_grid, lambda_grid, hstar)
z_max <- max(density, na.rm=TRUE)
persp(
  theta_grid, 
  lambda_grid, 
  density, 
  zlim=c(0, 1.1*z_max),
  xlab="theta",
  ylab="lambda",
  theta = 135, 
  phi = 30, 
  expand=0.5,
  ticktype="detailed")
```

@@

```{r}
density <- density / sum(density, na.rm=TRUE)
density %>%
  as.vector %>%
  sort %>%
  rev -> sorted_density
cumulative_p <- 0
for (p in sorted_density) {
  cumulative_p <- p + cumulative_p
  density[density==p] <- cumulative_p
}
contour(
  theta_grid, 
  lambda_grid, 
  density,
  labcex=1,
  levels=c(10, 25, 50, 75, 90)/100,
  xlab="theta",
  ylab="lambda")
```

You can see that the hedging hyperprior did its job. The most probable values of $\lambda$ are small and the prior mean (`r b0/(a0-1)`) is largely ignored in favor of the mean of the data (`r xbar`).

### 2023-03-09 text/23/normal-hedging-hyperprior.Rmd
Converted normal-hedging-hyperprior to use precisions

@@

$g(\mu; \mu_0, \sigma_0)=\frac{1}{\sigma_0 \sqrt{2 \pi}}e^{-(\mu-\mu_0)^2 / 2\sigma_0^2}$

Simplify this by removing any constants not involving $\mu$.

$g(\mu; \mu_0, \sigma_0) \propto e^{-(\mu-\mu_0)^2 / 2\sigma_0^2}$

and you collect a sample of n data points, $\mathbf{X} = X_1, X_2, ..., X_n$. This gives you a likelihood of

$L(\mathbf{X} | \mu) = \frac{1}{(\sigma \sqrt{2 \pi})^n}e^{-\sum(X_i-\mu)^2 / 2\sigma^2}$

@@

$L(\mathbf{X} | \mu)= \frac{1}{(\sigma \sqrt{2 \pi})^n}e^{-\sum(X_i-\bar{X}+\bar{X}-\mu)^2 / 2\sigma^2}$

$L(\mathbf{X} | \mu) = \frac{1}{(\sigma \sqrt{2 \pi})^n}e^{-((n-1)S^2 / 2\sigma^2}e^{-\sum(\bar{X}-\mu)^2) / 2\sigma^2}$

Pull out anything that is constant with respect to $X_i$ or $\mu$.

$L(\mathbf{X}| \mu) \propto e^{-\sum(\bar{X}-\mu)^2) / 2\sigma^2}$

@@

$h(\mu | \mathbf{X}) \propto e^{-(\mu-\mu_0)^2 / 2\sigma_0^2}e^{-n(\bar{X}-\mu)^2) / 2\sigma^2}$

@@

$h(\mu | \mathbf{X}) \propto e^{-((\frac{\sigma}{\sigma_0}(\mu_0-\mu))^2 + n (\bar{X}-\mu)^2) / 2\sigma^2}$

which, thanks to the miracle of conjugancy, is an inverse gamma distribution. The posterior mean

$\frac{\beta + n\bar X}{\alpha-1+n}$

@@

$\Big(\frac{\alpha-1}{\alpha-1+n}\Big)\frac{\beta}{\alpha-1}+\Big(\frac{n}{\alpha-1+n}\Big)\bar X$

>>>>

The mathematical details are a bit easier to follow if you let

$\tau = \frac{1}{\sigma}$

The parameter $\tau$ is called the precision. The normal density function in terms of $\tau$ is

$f(X) = \frac{\tau}{\sqrt{2 \pi}}e^{- \tau^2 (X-\mu)^2}$

Simplify this by removing any constants not involving $X$, $\mu$, or $\tau$.

$f(X) \propto \tau e^{- \tau^2 (X-\mu)^2}$

@@

$g(\mu; \mu_0, \tau_0) \propto \tau_0 e^{-\tau_0^2 (\mu-\mu_0)^2}$

You have collected a sample of n data points, $\mathbf{X} = X_1, X_2, ..., X_n$. This gives you a likelihood of

$L(\mathbf{X} | \mu) \propto \tau^n e^{-\tau^2\sum(X_i-\mu)^2}$

@@

$L(\mathbf{X} | \mu) \propto \tau^n e^{-\sum \tau (X_i-\bar{X}+\bar{X}-\mu)^2}$

When you square the term inside the sum, the cross-product cancels out, leaving you with

$L(\mathbf{X} | \mu) \propto \tau^n e^{-\tau^2(\sum(X_i-\bar{X})^2} e^{-\tau^2 n (\bar{X}-\mu)^2}$

$L(\mathbf{X} | \mu) \propto \tau^n e^{-\tau^2((n-1)S^2} e^{-\tau^2 n (\bar{X}-\mu)^2}$

@@

$h(\mu | \mathbf{X}) \propto \tau_0 e^{-\tau_0^2(\mu-\mu_0)^2} \tau^n e^{-\tau((n-1)S^2} e^{-\tau^2\sum(\bar{X}-\mu)^2)}$

@@

$h(\mu | \mathbf{X}) \propto $

which, thanks to the miracle of conjugancy, is a normal distribution. The posterior mean

$\frac{\tau n\bar{X}\beta + \tau_0 \mu_0}{n\tau+\tau0}$

@@

The posterior precision is

$\tau0+n\tau$
